Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2016-10-03 article.tex
theory correction

Numerical optimization is a diverse field with as many algorithms as there are optimization problems around. For our application, derivative-based methods and in particular the Newton method plays a major role.

@@

Parameter estimates $\hat \theta$  are obtained by non-linear optimization of the log-likelihood function $l(\theta)$. Although being a function of the parameters,  the log-likelihood depends on the data, too. Consequently, "equivalent" random ralizations of the data would lead to different parameter estimates $\hat\theta$. Given $\hat\theta$ for one random realization of the data, the question is, if we can reject other parameters $\theta$ with a certain confidence level. This leads to the related question of parameter convidence intervals.

A useful tool to derive confidence intervals is the profile likelihood. Consider a parameter of interest, $\theta_i$, being one of the parameters in the vector $\theta$. Furthermore, let $g_x(\theta) = \theta_i - x = 0$ be a parameter constraint. Then the function 
\begin{align}
	\begin{aligned}
		{\rm pl}_i:\quad \mathbb R & \longrightarrow \mathbb R\\
		x & \longmapsto \min_{(\theta, \lambda)} \big[ l(\theta) + \lambda g_x(\theta) \big]
	\end{aligned}
	\label{eq:pl}
\end{align}
is called the profile likelihood of the $i^{\;\rm th}$ paramameter and the path $x \mapsto (\hat\theta_x, \hat\lambda_x)$ along the minimizing parameters is called the profile likelihood path. Hence, the profile likelihood ${\rm pl}_i$  returns the minimal log-likelihood under the constraint $\theta_i \stackrel{!}{=} x$. By construction, the inequality $D := {\rm pl}_i(x) - {\rm pl}_i(\hat \theta_i) \geq 0$ holds for all $x$, assuming equality for $x = \hat\theta_i$. Interpreting the model with free $\theta$ as the null model and that with $\theta_i$ fixed to $x$ as the alternative model, the value $D$ is twice the log of the likelihood-ratio between those. Hence, a likelihood-ratio test can be performed to accept or reject the alternative model at a given confidence level. To compute the corresponding threshold, we assume Wilks' theorem according to which the thresholds are the quantiles of the $\chi^2$ distribution with one degree of freedom. Consequently, the $1\sigma, 2\sigma, \dots, n\sigma$-confidence intervals of $\theta_i$ are those values of $x$ for which ${\rm pl}_i(x) \leq 1, 2, \dots, n$, respectively.

The reason to formulate the profile likelihood by Lagrangian multipliers is that we can directly use eq.~\eqref{eq:pl} to derive the profile likelihood path $(\hat \theta_x, \hat\lambda_x)$ with respect to $x$:
\begin{align}
	\frac{\rm d}{{\rm d} x}
	\begin{pmatrix}
		\hat\theta_x \\ \hat\lambda_x
	\end{pmatrix} =
	\begin{pmatrix}
		Hl(\hat\theta_x) & \nabla g_x(\hat \theta_x)\\
		\nabla g_x(\hat\theta_x)^T & 0
	\end{pmatrix}^{-1}
	\begin{pmatrix}
		0 \\ 1
	\end{pmatrix}.
	\label{eq:path}
\end{align}
Eq.~\eqref{eq:path} forms the basis of what is implemented in \pkg{dMod} for computing the profile likelihood.

>>>>

Numerical optimization is a diverse field with as many algorithms as there are optimization problems around. For our application, derivative-based methods, and in particular the Newton method, play a major role.

@@

Parameter estimates $\hat \theta$  are obtained by non-linear optimization of the log-likelihood function $l(\theta)$. Although being a function of the parameters,  the log-likelihood depends on the data, too. Consequently, "equivalent" random realizations of the data would lead to different parameter estimates $\hat\theta$. Given $\hat\theta$ for one random realization of the data, the question is, if we can reject other parameters $\theta$ with a certain confidence level. This leads to the related question of parameter convidence intervals.

A useful tool to derive confidence intervals is the profile likelihood. Consider a parameter of interest, $\theta_i$, being one of the parameters in the vector $\theta$. Furthermore, let $c_{\tau}(\theta) = \theta_i - \tau = 0$ be a parameter constraint. 
Then, extending the likelihood with the constraint via a Lagrange multiplier $\lambda$ yields
\begin{align}
	\begin{aligned}
		{\rm pl}_i:\quad \mathbb R & \longrightarrow \mathbb R\\
		\tau & \longmapsto \min_{(\theta, \lambda)} \big[ l(\theta) + \lambda c_{\tau}(\theta) \big]
	\end{aligned}
	\label{eq:pl}
\end{align}
which called the profile likelihood of the $i^{\;\rm th}$ parameter and the path $\tau \mapsto (\hat\theta_{\tau}, \hat\lambda_{\tau})$ along the minimizing parameters is called the profile likelihood path. Hence, the profile likelihood ${\rm pl}_i$ returns the minimal log-likelihood under the constraint $\theta_i \stackrel{!}{=} \tau$. By construction, the inequality $D := {\rm pl}_i(\tau) - {\rm pl}_i(\hat \theta_i) \geq 0$ holds for all $\tau$, assuming equality for $\tau = \hat\theta_i$. Interpreting the model with free $\theta$ as the null model and that with $\theta_i$ fixed to $\tau$ as the alternative model, the value $D$ is twice the log of the likelihood-ratio between those. Hence, a likelihood-ratio test can be performed to accept or reject the alternative model at a given confidence level. To compute the corresponding threshold, we assume Wilks' theorem according to which the thresholds are the quantiles of the $\chi^2$ distribution with one degree of freedom. Consequently, the $1\sigma, 2\sigma, \dots, n\sigma$-confidence intervals of $\theta_i$ are those values of $\tau$ for which ${\rm pl}_i(\tau) \leq 1, 2, \dots, n$, respectively.

The reason to formulate the profile likelihood by Lagrangian multipliers is that we can directly use eq.~\eqref{eq:pl} to derive the profile likelihood path $(\hat \theta_{\tau}, \hat\lambda_{\tau})$ with respect to $\tau$:
\begin{align} 
	\frac{\rm d}{{\rm d} \tau}
	\begin{pmatrix}
		\hat\theta_{\tau} \\ \hat\lambda_{\tau}
	\end{pmatrix} =
	\begin{pmatrix}
		Hl(\hat\theta_{\tau}) & \nabla c_{\tau}(\hat \theta_{\tau})\\
		\nabla c_{\tau}(\hat\theta_{\tau})^T & 0
	\end{pmatrix}^{-1}
	\begin{pmatrix}
		0 \\ 1
	\end{pmatrix}.
	\label{eq:path}
\end{align}
Eq.~\eqref{eq:path} forms the basis of what is implemented in \pkg{dMod} for computing the profile likelihood.
